Automatic Matching of Bullet Lands

### 2015-12-26 imaging-paper.Rnw
alignment in table

% put your definitions there:
\startlocaldefs
\usepackage{enumitem}
\usepackage[hang, small,labelfont=bf,up,textfont=it,up]{caption} % Custom captions under/above floats in tables or figures
\usepackage{subcaption}
\usepackage{lineno,hyperref}
\modulolinenumbers[5]

@@

my.loess$fitted + ylab("Surface measurements")
@
\end{subfigure}
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{\label{fig:loess_resid} Residuals of loess fit for bullet~1-5.
This defines the {\it signature} of bullet~1-5. Blue areas show  95\% bootstrap confidence intervals (based on $B=500$ bootstrap samples).
}
%<<loess_resid, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\textwidth', warning=FALSE>>=
%my.loess$resid + ylab("Residuals of loess Fit") 
%@
<<boot_loess, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\textwidth', warning=FALSE>>=
boot.loess <- boot_fit_loess(br111_bullet, br111.groove)
poly <- with(boot.loess, data.frame(x = c(y,rev(y)), y=c(high, rev(low))))
qplot(x=x, y=y, geom="polygon", fill=I("steelblue"), data=poly) +
  geom_line(aes(x=y, y=resid), data=my.loess$data) +
  ylab("Residuals of loess fit") + xlab("y") + theme_bw()
@
\end{subfigure}
\caption{Fit and residuals of a loess fit to bullet~1-5 (Barrel~1).}
\end{figure}
%
Figure~\ref{fig:loessfits} shows the signatures of all lands of bullet 2 (in grey) fired from the same barrel as bullet~1. The darker line overlaid on each of the signatures, shows signature from bullet~1-5. 
%We can reproduce a similar figure to the one from the circular fits by overlaying the residuals from each loess fit for bullet 2 onto an image of the loess residuals from bullet~1-5. This is shown in Figure~\ref{fig:loessfits}. 
Bullet 2-1 is the best match for bullet~1-5, but the alignment is not quite  correct.

@@

Using recursive partitioning, we fit a decision tree~\citep{breiman:1984, rpart, rpart.plot} to predict matches between lands based on features derived from the image files. The resulting tree is shown in Figure~\ref{fig:tree}. A total of \Sexpr{sum(bstats$match[bstats$pred>0.5])} lands is being matched correctly. Interestingly, the number of consecutive matching striae does not feature in this evaluation. 
Instead of CMS, cross-correlation (ccf) between the cross sections is very important in the matching process  by the decision tree. Aside from  cross correlation, the number of matches is involved in the decision. 
Between cross-correlation and CMS, cross-correlation has higher  predictive power. This  does not  contradict earlier findings emphasizing the value of CMS on visual assessments of bullet matches: in those papers, assessments were based on purely visual inspection of either actual bullets or 2d microscopic images of bullets.
Neither one of these methods allows for an assessment of cross-correlations. This is one of the benefits of switching to a digitalized version of the images that preserves the 3d surface structure. \hh{The emphasis on cross-correlation is similar to the results of the study by \citet{ma:2004}. However, in this study authors did not consider the number of matches and non-matches.}

@@

Another benefit of the digitized version of the images is that we can not only apply a single decision tree, but several hundreds of them to combine in a random forest~\citep{breiman:2001, randomForest}.  For each of the trees in a random forest, only two thirds of the observations are used for fitting, while  the remaining third  are used to evaluate the tree's predictive power and accuracy, or its reverse, the error rate. Because errors are determined from the one third of held-back observations, this error rate is called the out-of bag (OOB) error. 
Figure~\ref{fig:oob} shows the cumulative out-of-bag error (OOB) rate for 300 trees. 
%
\begin{figure}[hbtp]
  \centering
<<oob, echo=FALSE, fig.width=7, fig.height=3, out.width='.5\\textwidth'>>=
qplot(id, OOB, geom="line", data=errors) + theme_bw() + ylim(c(0,NA)) + xlab("Number of trees") + ylab("Out of Bag Error (OOB)")
@
\caption{\label{fig:oob}Cumulative out-of-bag error rate of a random forest fit to predict land-to-land matches from image features.}
\end{figure}
%
After about 100 trees, the error rate of land-to-land comparisons stabilizes at 0.0038. This is a weighted average between a zero false positive error rate and an error rate of false negatives of 0.23. \hh{This out-of-bag error rate is over-estimating the actual error in the Hamby study: here, the final random forest based on 300 trees is able to correctly predict all known  matches and non-matches (see Figures~\ref{fig:treeresult} and ~\ref{fig:tree-forest}).}
Note that this error rate is based on land-to-land comparisons and is much lower for bullet-to-bullet matches. In the case of the Hamby study, even for the single tree the overall error rate is zero, if we require for a bullet match that at least two of a bullet's lands are matched. % (see Figure~\ref{fig:treeresult}). 
%' \begin{figure}[hbtp]
%'   \centering
%' <<overall-tree, echo=FALSE, fig.width=7, fig.height=4, out.width='.5\\textwidth'>>=
%' bstats$bullet <- gsub("-[0-9]$", "", bstats$b2)
%' bullets <- bstats %>% filter(match, !flagged) %>% group_by(bullet) %>% summarize(
%'   n = n(),
%'   pred = sum(pred > 0.5),
%'   forest = sum(forest > 0.5)
%' )
%' library(reshape2)
%' bm <- melt(data=bullets, id.var=c("bullet", "n"))
%' 
%' bm$prediction <- c("Tree", "Random Forest")[as.numeric(bm$variable)]
%' qplot(data=bm, 100*value/n, reorder(bullet, value/n), size=I(2.5), 
%'       shape=prediction, colour=prediction, alpha=I(0.6)) + 
%'   theme_bw()  + ylab("") + 
%' #  scale_shape_manual(values=c(1,2)) +
%'   scale_x_continuous("Percent of correctly predicted land-to-land matches", 
%'                      breaks = c(0,25,50,75,100), limits=c(0,100)) +
%'   theme(legend.position = "bottom")
%' @
%' \caption{\label{fig:treeresult}Overall predictions for each bullet based on the decision tree (maximum possible number is 12, except for Q and B, where the maximum is 11). Both results for the tree and the random forest are shown. In the tree, at least \hh{40\% of all possible} lands of a bullet are  matched correctly, while the random forest achieves perfect results. Unknown bullet B is the hardest for the tree to match.}
%' \end{figure}
%' %
\hh{This makes the errors in the automated approach smaller than the human error in the Hamby study. Out of the 507 participants, who returned results, eight (out of $15 \times 507 = 7,605$) bullets were not matched conclusively, resulting in a false negative rate of 0.0011.}

@@

Besides a probabilistic quantification of matches, we get from a random forest an assessment of the importance of each of the features derived from the bullets' 3d topological surface measurements. Table~\ref{tab:importance} shows an overview of the importance of each variable measured as the mean decrease in the gini index when the variable in question is included in a tree.
\begin{table}[tbhp]
\caption{\label{tab:importance}Table of  features derived from bullet image ordered by importance in predicting matches. Importance is measured in terms of mean decrease in gini index when including the variable in a decision tree. Match and Non-Match show averages of each of the variables for matches and non-matches only. In parentheses behind the means, standard deviations are given}
\centering
<<importance, echo=FALSE, results='asis'>>=
imp <- data.frame(importance(rtrees))
imp$Variable <- row.names(imp)
imp <- imp[order(-imp$MeanDecreaseGini),]
names(imp)[1] <- "Importance"
imp <- imp[,2:1]
imp$meanMatch <- sapply(imp$Variable, function(var) {
  mean(subset(bstats, match)[,var])
})
imp$sdMatch <- sapply(imp$Variable, function(var) {
  sd(subset(bstats, match)[,var])
})
imp$meanNonMatch <- sapply(imp$Variable, function(var) {
  mean(subset(bstats, !match)[,var])
})
imp$sdNonMatch <- sapply(imp$Variable, function(var) {
  sd(subset(bstats, !match)[,var])
})
imp$meanMatch <- round(imp$meanMatch,1)
imp$sdMatch <- sprintf("(%5.2f)", imp$sdMatch)
imp$meanNonMatch <- round(imp$meanNonMatch,1)
imp$sdNonMatch <- sprintf("(%7.2f)", imp$sdNonMatch)
row.names(imp) <- 1:nrow(imp)
imp$Variable[c(2,4)] <- c("#matches", "#non-matches")
names(imp)[-(1:2)] <- c("Match", "(sd)", "Non-Match", "(sd)")
print(xtable(imp, align="clrrcrc"), floating = FALSE)
@
\end{table}

>>>>

% put your definitions there:
\startlocaldefs
\usepackage{enumitem}
\usepackage[hang, small,labelfont=bf,up,textfont=it,up]{caption} % Custom captions under/above floats in tables or figures
\usepackage{subcaption}
\usepackage{lineno,hyperref}
\modulolinenumbers[5]
\usepackage{dcolumn}
\newcolumntype{d}[1]{D{.}{.}{#1}}

@@

my.loess$fitted + ylab("Surface measurements")
@
\end{subfigure}
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{\label{fig:loess_resid} Residuals of loess fit for bullet~1-5.
This defines the {\it signature} of bullet~1-5. Blue areas show  95\% bootstrap confidence intervals (based on $B=1,000$ bootstrap samples).
}
%<<loess_resid, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\textwidth', warning=FALSE>>=
%my.loess$resid + ylab("Residuals of loess Fit") 
%@
<<boot_loess, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\textwidth', warning=FALSE>>=
boot.loess <- boot_fit_loess(br111_bullet, br111.groove)
poly <- with(boot.loess, data.frame(x = c(y,rev(y)), y=c(high, rev(low))))
qplot(x=x, y=y, geom="polygon", fill=I("steelblue"), data=poly) +
  geom_line(aes(x=y, y=resid), data=my.loess$data) +
  ylab("Residuals of loess fit") + xlab("y") + theme_bw()
@
\end{subfigure}
\caption{Fit and residuals of a loess fit to bullet~1-5 (Barrel~1).}
\end{figure}
%
Figure~\ref{fig:loessfits} shows the signatures of all lands of bullet 2 (in grey) fired from the same barrel as bullet~1. The darker line overlaid on each of the signatures, shows signature from bullet~1-5. 
%We can reproduce a similar figure to the one from the circular fits by overlaying the residuals from each loess fit for bullet 2 onto an image of the loess residuals from bullet~1-5. This is shown in Figure~\ref{fig:loessfits}. 
Bullet 2-1 is the best match for bullet~1-5, but the alignment is not quite  correct.

@@

Using recursive partitioning, we fit a decision tree~\citep{breiman:1984, rpart, rpart.plot} to predict matches between lands based on features derived from the image files. The resulting tree is shown in Figure~\ref{fig:tree}. A total of \Sexpr{sum(bstats$match[bstats$pred>0.5])} lands is being matched correctly. Interestingly, the number of consecutive matching striae does not feature in this evaluation. 
Instead of CMS, cross-correlation (ccf) between the signatures is very important in the matching process  by the decision tree. Aside from  cross correlation, the number of matches is involved in the decision. 
Between cross-correlation and CMS, cross-correlation has higher  predictive power. This  does not  contradict earlier findings emphasizing the value of CMS on visual assessments of bullet matches: in those papers, assessments were based on purely visual inspection of either actual bullets or 2d microscopic images of bullets.
Neither one of these methods allows for an assessment of cross-correlations. This is one of the benefits of switching to a digitalized version of the images that preserves the 3d surface structure. \hh{The emphasis on cross-correlation is similar to the results of the study by \citet{ma:2004}. However, in this study authors did not consider the number of matches and non-matches.}

@@

Another benefit of the digitized version of the images is that we can not only apply a single decision tree, but several hundreds of them to combine in a random forest~\citep{breiman:2001, randomForest}.  For each of the trees in a random forest, only two thirds of the observations are used for fitting, while  the remaining third  are used to evaluate the tree's predictive power and accuracy, or its reverse, the error rate. Because errors are determined from the one third of held-back observations, this error rate is called the out-of bag (OOB) error. 
Figure~\ref{fig:oob} shows the cumulative out-of-bag error (OOB) rate for 300 trees. 
%
\begin{figure}[hbtp]
  \centering
<<oob, echo=FALSE, fig.width=7, fig.height=3, out.width='.5\\textwidth'>>=
qplot(id, OOB, geom="line", data=errors) + theme_bw() + ylim(c(0,NA)) + xlab("Number of trees") + ylab("Out of Bag Error (OOB)")
@
\caption{\label{fig:oob}Cumulative out-of-bag error rate of a random forest fit to predict land-to-land matches from image features.}
\end{figure}
%
After about 100 trees, the error rate of land-to-land comparisons stabilizes at 0.0038. This is a weighted average between a zero false positive error rate and an error rate of false negatives of 0.23. \hh{This out-of-bag error rate is over-estimating the actual error in the Hamby study: here, the final random forest based on 300 trees is able to correctly predict all known  matches and non-matches (see Figure~\ref{fig:tree-forest}).}
Note that this error rate is based on land-to-land comparisons and is much lower for bullet-to-bullet matches. In the case of the Hamby study, even for the single tree the overall error rate is zero, if we require for a bullet match that at least two of a bullet's lands are matched. % (see Figure~\ref{fig:treeresult}). 
%' \begin{figure}[hbtp]
%'   \centering
%' <<overall-tree, echo=FALSE, fig.width=7, fig.height=4, out.width='.5\\textwidth'>>=
%' bstats$bullet <- gsub("-[0-9]$", "", bstats$b2)
%' bullets <- bstats %>% filter(match, !flagged) %>% group_by(bullet) %>% summarize(
%'   n = n(),
%'   pred = sum(pred > 0.5),
%'   forest = sum(forest > 0.5)
%' )
%' library(reshape2)
%' bm <- melt(data=bullets, id.var=c("bullet", "n"))
%' 
%' bm$prediction <- c("Tree", "Random Forest")[as.numeric(bm$variable)]
%' qplot(data=bm, 100*value/n, reorder(bullet, value/n), size=I(2.5), 
%'       shape=prediction, colour=prediction, alpha=I(0.6)) + 
%'   theme_bw()  + ylab("") + 
%' #  scale_shape_manual(values=c(1,2)) +
%'   scale_x_continuous("Percent of correctly predicted land-to-land matches", 
%'                      breaks = c(0,25,50,75,100), limits=c(0,100)) +
%'   theme(legend.position = "bottom")
%' @
%' \caption{\label{fig:treeresult}Overall predictions for each bullet based on the decision tree (maximum possible number is 12, except for Q and B, where the maximum is 11). Both results for the tree and the random forest are shown. In the tree, at least \hh{40\% of all possible} lands of a bullet are  matched correctly, while the random forest achieves perfect results. Unknown bullet B is the hardest for the tree to match.}
%' \end{figure}
%' %
\hh{This makes the errors in the automated approach smaller than the human error in the Hamby study. Out of the 507 participants, who returned results, eight (out of $15 \times 507 = 7,605$) bullets were not matched conclusively, resulting in a false negative rate of 0.0011.}

@@

Besides a probabilistic quantification of matches, we get from a random forest an assessment of the importance of each of the features derived from the bullets' 3d topological surface measurements. Table~\ref{tab:importance} shows an overview of the importance of each variable measured as the mean decrease in the gini index when the variable in question is included in a tree.
\begin{table}[tbhp]
\caption{\label{tab:importance}Table of  features derived from bullet image ordered by importance in predicting matches. Importance is measured in terms of mean decrease in gini index when including the variable in a decision tree. Averages (and standard deviations) for known matches (KM) and known non-matches (KNM) are shown in the last four columns.}
\centering
<<importance, echo=FALSE, results='asis', warning=FALSE>>=
imp <- data.frame(importance(rtrees))
imp$Variable <- row.names(imp)
imp <- imp[order(-imp$MeanDecreaseGini),]
names(imp)[1] <- "Importance"
imp <- imp[,2:1]
imp$meanMatch <- sapply(imp$Variable, function(var) {
  mean(subset(bstats, match)[,var])
})
imp$sdMatch <- sapply(imp$Variable, function(var) {
  sd(subset(bstats, match)[,var])
})
imp$meanNonMatch <- sapply(imp$Variable, function(var) {
  mean(subset(bstats, !match)[,var])
})
imp$sdNonMatch <- sapply(imp$Variable, function(var) {
  sd(subset(bstats, !match)[,var])
})
imp$meanMatch <- round(imp$meanMatch,1)
imp$sdMatch <- sprintf("(%5.2f)", imp$sdMatch)
imp$meanNonMatch <- round(imp$meanNonMatch,1)
imp$sdNonMatch <- sprintf("(%7.2f)", imp$sdNonMatch)
row.names(imp) <- 1:nrow(imp)
imp$Variable[c(2,4)] <- c("#matches", "#non-matches")
names(imp)[-(1:2)] <- c("KM", "(sd)", "KNM", "(sd)")
print(xtable(imp, align="clrrd{3.2}rd{3.2}", digits=c(0,0,1,1,2,1,2)), 
      floating = FALSE)
@
\end{table}
